In-silico molecular enrichment and clearance of the human intracranial space

### 2025-06-20 paper/main.tex
Spell-check and PVSAS

% What are our research questions and how do we answer them
In this study, by leveraging geometric model reduction and
mixed-dimensional modelling~\cite{masri2024modelling}, structural
magnetic resonance (MR) images~\cite{hodneland2019new}, and high
performance computing, we introduce an integrated computational model
of intracranial molecular transport. Focusing on the
interplay between perivascular pathways, pulsatility and CSF flow
dynamics, the model predicts the temporal evolution and spatial
distribution of a solute concentration within a detailed geometric
representation of the human SAS and ventricular system, networks of
surface PVSs and the brain parenchyma. In terms of transport dynamics,
we account for heterogeneous diffusion, dispersive mixing induced by
cardiac and respiratory pulsatility in the CSF spaces and PVSs,
convective fluid flow driven by CSF production and peristaltic
pumping, as well as solute exchange and clearance across
semi-permeable membranes. Comparing with glymphatic MRI
studies~\cite{ringstad2017glymphatic, ringstad2018brain,
  watts2019measuring}, the in-silico predictions accurately represent
molecular enrichment patterns, timing and intercompartmental
distributions. This open platform~\cite{ZENODO} thus provides a
technological opportunity for qualitatively and quantitatively
exploring key open questions relating to molecular movement within the
human brain environment such as the role of pulsatility, perivascular
pathways, structural compartmentalization, or morphology.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure3a.png}
    \caption{\textbf{A static pressure gradient and peristaltic pumping are potential drivers of perivascular flow.}
    A) Surface arterial network colored by arterial radius with vessel locations labeled;
    B) Net PVS flow induced by CSF production alone (see also ~\Cref{fig:csf}D).
    Histograms show the relative frequencies of local periarterial velocities. Positive values indicate antegrade PVS flow, while negative values indicate retrograde PVS flow;
    C) Illustration of the concept of peristaltic pumping;
    D) histograms of estimated net PVS flow induced by pulse wave or vasomotion peristaltic pumping;
    E) Estimated net PVS flow induced by pulse wave peristaltic pumping alone.
    F) Estimated net PVS flow induced by vasomotion/slow wave peristaltic pumping alone.}
    \label{fig:pvs_a}
\end{figure}

@@

Our simulation results highlight that reproducing clinically observed tracer timings requires flow drivers beyond CSF production and cardiac peristalsis.
Similar to the CSF spaces, dispersion due to cardiac and respiratory pulsation has been hypothesized to enhance transport in the perivascular spaces. However, recent estimates indicate that the effect of dispersion does not exceed a factor of 2 compared to diffusion alone \cite{asgari2016glymphatic,sharp2019dispersion,bojarskaite2023sleep,asgari2016glymphatic,troyetsky2021dispersion}. This conclusion is supported by our own model, where a 100-fold increase in the PVS dispersion coefficient did not substantially accelerate transport (\cref{fig:pvs_b}H). 
To illustrate the need for a stronger advective driver, we compared our findings to tracer arrival times reported by Eide and Ringstad~\cite{eide2024functional}, who observed a delay of just 15.3 min between the M1 and M2 segments of the MCA. Our baseline model, accounting only for CSF production and cardiac-driven flow, predicted a physiologically unrealistic delay of over three hours. In sharp contrast, our 'high PVS flow' scenario, which incorporates the effects of low-frequency vasomotion, yielded a much more consistent delay of 24 minutes. Furthermore, after adjusting for the 13 min travel time from injection to the craniocervical junction~\cite{eide2024functional}, the 'high PVS flow' model's FTAs (84––108 min for M1) aligns well with the upper end of the reported clinical range (37.8 $\pm$ 47.0), particularly when considering potentially delayed transport due to the limited caudal extend of our PVS network.
Therefore, we interpret these results as strong indication that PVS flow driven by CSF production and cardiac peristalsis alone is insufficient to explain rapid perivascular transport observed in humans.

@@

% Variations with sleep, exercise
Lifestyle factors such as sleep~\cite{xie2013sleep, miao2024brain,
  bojarskaite2023sleep, eide2021sleep, vinje2023human,
  hauglund2025norepinephrine, larsen2024sleep},
exercise~\cite{holstein2018voluntary, yildiz2022immediate}, and
alcohol intake~\cite{lundgaard2018beneficial} affect molecular
enrichment and clearance of the brain. In particular, pial arteries
display higher-amplitude low-frequency vasomotion during NREM and
intermediate sleep states, while REM sleep is associated with reduced
PVS width~\cite{bojarskaite2023sleep}. Peristaltic pumping is more
effective under both of these configurations, resulting in higher
estimated net PVS flow velocities~\cite{gjerde2023directional}. Even
higher PVS velocities would amplify the PVS flow effects simulated
here, with earlier arrival times along the PVSs, clear enrichment in
adjacent spaces, and accelerated molecular enrichment
overall. Moreover, sleep is linked with low-frequency oscillations in
human CSF flow~\cite{fultz2019coupled}. As we have shown here, the
level of dispersion in the CSF spaces strongly shapes molecular
enrichment patterns. While we, at baseline, account for the dispersion
induced by cardiac and respiratory pulsatility, the high dispersion
scenarios considered can be viewed as representative of additional
pulsatility induced by e.g.~neural waves~\cite{fultz2019coupled,
  williams2023neural} or other respiration patterns depending on
activity level. More broadly, we consider this as a key strength of
the current in-silico framework: allowing for integrating and
exploring the mechanistic macroscale implications of changes in
pulsatility at smaller spatial or temporal scales.
This approach of modeling pulsatility through an effective dispersion 
factor is a simplification that does not capture the full complexity
of transient fluid dynamics, such as non-uniform wall motion and ependymal
cilia within the ventricular system~\cite{siyahhan2014flow}, dynamic CSF
flow at the craniocervical junction, or the effect of microanatomical
features such as subarachnoid trabeculae on flow resitance and dispersion
in the SAS. However, our approach is supported by recent clinical 
observations. High-resolution MRI measurements by Hirschler et
al.~\cite{hirschler2024region} revealed that CSF mobility in regions
with high pulsatility is approximately tenfold greater than the 
self-diffusion of water, an enhancement factor consistent with 
the values computed in our model.

@@

We give a concise summary of the numerical approach here (see
Supplementary methods for a comprehensive description). To
solve~\eqref{eq:multi_transport} numerically, we use the interior
penalty discontinuous Galerkin (DG) method with weighted averages and
upwinding for the convection term in the 3D
domains~\cite{ern2009discontinuous} -- to ensure numerical stability
and minimize the artificial presence of negative concentration
values. For the 1D domain, we use continuous finite element spaces
defined over each of the networks and stabilized with numerical
diffusion when $\hat{u}$ is non--zero. Note that the 3D--1D coupling
in~\eqref{eq:multi_transport} leads to a 3D solution $c$ with low
regularity properties. Thus, one can only expect lower-order error
convergence near $\Lambda_a$ and $\Lambda_v$, see the references
\cite{masri2024discontinuous} and \cite{laurino2019derivation} for the
numerical analysis of DG and continuous Galerkin schemes,
respectively. However, away from these 1D networks, almost optimal
approximation is
expected~\cite{masri2023discontinuous,koppl2016local}. To numerically
solve the Stokes equations to compute the CSF flow velocity and
pressure in the ventricular system and SAS, we use finite element
spaces that preserve the incompressibility condition on the discrete
level~\cite{hong2016robust}. This requirement is important for
numerical stability when subsequently
solving~\eqref{eq:multi_transport}\cite{cesmelioglu2022compatible}.
The spaces we used for the velocity are continuous along the normal
direction; continuity of the tangential component is enforced via
interior penalty approaches~\cite{hong2016robust}. Optimal
approximation properties are expected for this scheme. These numerical
methods were implemented using the FEniCS finite element
framework~\cite{alnaes2015fenics}, and the 3D-1D coupling is handled
by the extension FEniCS\textsubscript{ii}~\cite{kuchta2020assembly}.
The correctness and accuracy of the numerical solutions were verified by a series of
numerical verification experiments (see Supplementary information S1.5).
The majority of the simulation time was spend on the Stokes flow and transport model simulations. Each Stokes flow simulation utilizes a hybrid parallel approach with 16 MPI processes and 4 threads each, requiring approximately 300 GB of RAM and completing in 23 minutes. The transport model is limited to serial execution by FEniCS\textsubscript{ii}, but employes 16 threads for the linear system factorization, taking 22 minutes and consuming about 50 GB of memory on a AMD EPYC 9684X machine.

>>>>

% What are our research questions and how do we answer them
In this study, by leveraging geometric model reduction and
mixed-dimensional modelling~\cite{masri2024modelling}, structural
magnetic resonance (MR) images~\cite{hodneland2019new}, and high
performance computing, we introduce an integrated computational model
of intracranial molecular transport. Focusing on the interplay between
perivascular pathways, pulsatility and CSF flow dynamics, the model
predicts the temporal evolution and spatial distribution of a solute
concentration within a detailed geometric representation of the human
SAS and ventricular system, in networks of surface PVSs (i.e.~those
surrounding arteries and veins in the SAS), and in the brain
parenchyma. In terms of transport dynamics, we account for
heterogeneous diffusion, dispersive mixing induced by cardiac and
respiratory pulsatility in the CSF spaces and PVSs, convective fluid
flow driven by CSF production and peristaltic pumping, as well as
solute exchange and clearance across semi-permeable
membranes. Comparing with glymphatic MRI
studies~\cite{ringstad2017glymphatic, ringstad2018brain,
  watts2019measuring, eide2024functional}, the in-silico predictions
accurately represent molecular enrichment patterns, timing and
intercompartmental distributions. This open platform~\cite{ZENODO}
thus provides a technological opportunity for qualitatively and
quantitatively exploring key open questions relating to molecular
movement within the human brain environment such as the role of
pulsatility, perivascular pathways, structural compartmentalization,
or morphology.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure3a.png}
    \caption{\textbf{A static pressure gradient and peristaltic pumping are potential drivers of perivascular flow.}
    A) Surface arterial network colored by arterial radius with vessel locations labeled;
    B) Net PVS flow induced by CSF production alone (see also~\Cref{fig:csf}D).
    Histograms show the relative frequencies of local periarterial velocities. Positive values indicate antegrade PVS flow, while negative values indicate retrograde PVS flow;
    C) Illustration of the concept of peristaltic pumping;
    D) histograms of estimated net PVS flow induced by pulse wave or vasomotion peristaltic pumping;
    E) Estimated net PVS flow induced by pulse wave peristaltic pumping alone.
    F) Estimated net PVS flow induced by vasomotion/slow wave peristaltic pumping alone.}
    \label{fig:pvs_a}
\end{figure}

@@

Importantly, our simulation results indicate that the tracer
enrichment observed clinically relies on the presence of one or more
flow drivers in addition to CSF production and cardiac
peristalsis. Dispersion due to cardiac and respiratory pulsations has
been hypothesized to enhance transport in the perivascular spaces as
well as in the CSF spaces. However, recent estimates indicate that the
effect of dispersion in the PVS does not exceed a factor of 2 compared
to diffusion alone
\cite{asgari2016glymphatic,sharp2019dispersion,bojarskaite2023sleep,asgari2016glymphatic,troyetsky2021dispersion}. Our
model demonstrates that not even a 100-fold increase in the PVS
dispersion coefficient accelerates transport substantially
(\Cref{fig:pvs_b}H). To illustrate the need for a stronger advective
driver, we compared our findings to tracer arrival times reported by
Eide and Ringstad~\cite{eide2024functional}, who observed a delay of
just 15.3 min between the M1 and M2 segments of the MCA. First, we
note that the spaces studied by Eide and
Ringstad~\cite{eide2024functional}, labeled therein as the PVSAS, are
naturally compared to the parts of our surface PVS networks embedded
within the SAS. Our baseline model, accounting only for CSF production
and cardiac-driven flow, predicted a physiologically unrealistic delay
of over three hours. In contrast, our high PVS flow scenario, which
incorporates the effects of low-frequency vasomotion, yielded a more
consistent delay of 24 minutes. Furthermore, after adjusting for the
13 min travel time from injection to the craniocervical
junction~\cite{eide2024functional}, the high PVS flow model's FTAs
(84--108 min for M1) aligns well with the upper end of the reported
clinical range (37.8 $\pm$ 47.0), particularly when considering
potentially delayed transport due to the limited caudal extent of our
PVS network. All in all, we interpret these results as strongly
indicative of that PVS flow driven by CSF production and cardiac
peristalsis alone is insufficient to explain the rapid perivascular
transport observed in humans.

@@

% Variations with sleep, exercise
Lifestyle factors such as sleep~\cite{xie2013sleep, miao2024brain,
  bojarskaite2023sleep, eide2021sleep, vinje2023human,
  hauglund2025norepinephrine, larsen2024sleep},
exercise~\cite{holstein2018voluntary, yildiz2022immediate}, and
alcohol intake~\cite{lundgaard2018beneficial} affect molecular
enrichment and clearance of the brain. In particular, pial arteries
display higher-amplitude low-frequency vasomotion during NREM and
intermediate sleep states, while REM sleep is associated with reduced
PVS width~\cite{bojarskaite2023sleep}. Peristaltic pumping is more
effective under both of these configurations, resulting in higher
estimated net PVS flow velocities~\cite{gjerde2023directional}. Even
higher PVS velocities would amplify the PVS flow effects simulated
here, with earlier arrival times along the PVSs, clear enrichment in
adjacent spaces, and accelerated molecular enrichment
overall. Moreover, sleep is linked with low-frequency oscillations in
human CSF flow~\cite{fultz2019coupled}. As we have shown here, the
level of dispersion in the CSF spaces strongly shapes molecular
enrichment patterns. While we, at baseline, account for the dispersion
induced by cardiac and respiratory pulsatility, the high dispersion
scenarios considered can be viewed as representative of additional
pulsatility induced by e.g.~neural waves~\cite{fultz2019coupled,
  williams2023neural} or other respiration patterns depending on
activity level. More broadly, we consider this as a key strength of
the current in-silico framework: allowing for integrating and
exploring the mechanistic macroscale implications of changes in
pulsatility at smaller spatial or temporal scales.
This approach of modeling pulsatility through an effective dispersion 
factor is a simplification that does not capture the full complexity
of transient fluid dynamics, such as non-uniform wall motion and ependymal
cilia within the ventricular system~\cite{siyahhan2014flow}, dynamic CSF
flow at the craniocervical junction, or the effect of microanatomical
features such as subarachnoid trabeculae on flow resistance and dispersion
in the SAS. However, our approach is supported by recent clinical 
observations. High-resolution MRI measurements by Hirschler et
al.~\cite{hirschler2024region} revealed that CSF mobility in regions
with high pulsatility is approximately tenfold greater than the 
self-diffusion of water, an enhancement factor consistent with 
the values computed in our model.

@@

We give a concise summary of the numerical approach here (see
Supplementary methods for a comprehensive description). To
solve~\eqref{eq:multi_transport} numerically, we use the interior
penalty discontinuous Galerkin (DG) method with weighted averages and
upwinding for the convection term in the 3D
domains~\cite{ern2009discontinuous} -- to ensure numerical stability
and minimize the artificial presence of negative concentration
values. For the 1D domain, we use continuous finite element spaces
defined over each of the networks and stabilized with numerical
diffusion when $\hat{u}$ is non--zero. Note that the 3D--1D coupling
in~\eqref{eq:multi_transport} leads to a 3D solution $c$ with low
regularity properties. Thus, one can only expect lower-order error
convergence near $\Lambda_a$ and $\Lambda_v$, see the references
\cite{masri2024discontinuous} and \cite{laurino2019derivation} for the
numerical analysis of DG and continuous Galerkin schemes,
respectively. However, away from these 1D networks, almost optimal
approximation is
expected~\cite{masri2023discontinuous,koppl2016local}. To numerically
solve the Stokes equations to compute the CSF flow velocity and
pressure in the ventricular system and SAS, we use finite element
spaces that preserve the incompressibility condition on the discrete
level~\cite{hong2016robust}. This requirement is important for
numerical stability when subsequently
solving~\eqref{eq:multi_transport}\cite{cesmelioglu2022compatible}.
The spaces we used for the velocity are continuous along the normal
direction; continuity of the tangential component is enforced via
interior penalty approaches~\cite{hong2016robust}. Optimal
approximation properties are expected for this scheme. These numerical
methods were implemented using the FEniCS finite element
framework~\cite{alnaes2015fenics}, and the 3D-1D coupling is handled
by the extension FEniCS\textsubscript{ii}~\cite{kuchta2020assembly}.
The correctness and accuracy of the numerical solutions were verified by a series of
numerical verification experiments (see Supplementary information S1.5).
The majority of the simulation time was spend on the Stokes flow and transport model simulations. Each Stokes flow simulation utilizes a hybrid parallel approach with 16 MPI processes and 4 threads each, requiring approximately 300 GB of RAM and completing in 23 minutes. The transport model is limited to serial execution by FEniCS\textsubscript{ii}, but employs 16 threads for the linear system factorization, taking 22 minutes and consuming about 50 GB of memory on a AMD EPYC 9684X machine.
